\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd} % \llbracket \rrbracket for the coalition bracket
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% proof-status tags: kernel-proved in i-orca, empirical, or open
\newcommand{\proved}{\textsc{[proved, i-orca]}}
\newcommand{\empirical}{\textsc{[empirical]}}
\newcommand{\open}{\textsc{[open]}}

\newcommand{\HH}{\mathcal{H}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\inner}[2]{\langle #1,\, #2\rangle}
\newcommand{\Uframe}{U}
\newcommand{\Gram}{G}

\title{\textbf{Projective Incidence Calculus}\\[0.3em]
\large A semiring-weighted incidence algebra for the transformer decode,\\
its frozen-model geometry, and its learnable frame}
\author{}
\date{Draft --- companion to \texttt{spec/PIC\_SPEC.md}}

\begin{document}
\maketitle

\begin{abstract}
We give an algebraic account of the transformer next-token decision as a \emph{semiring-weighted
incidence calculus over a continuous inner-product geometry}---\emph{Projective Incidence Calculus}
(PIC). Propositions (tokens) are points $U_v$ in a residual space $\HH=\RR^d$; sources (per-layer
contributions, or learned rules) are vectors $d_j$; the incidence of a source on a proposition is the
inner product $\inner{d_j}{U_v}$; and the decision aggregates incidences through a commutative semiring
selected by a temperature $T$. At $T=1$ the semiring is the log-semiring and PIC \emph{is} the forward
pass (softmax over summed logits); as $T\to0$ it is the tropical (max-plus) semiring and PIC is the
geometry of margins, Laguerre power diagrams, and a head/tail decode certificate, the two ends joined
by Maslov dequantization. PIC separates the \emph{frame} (intrinsic geometry of $\{U_v\}$: Gram
coherence, Welch floor, packing capacity) from the \emph{decode} (margins, participation ratio,
multiplicity, evaluated in the chosen semiring). This separation organizes a three-part program:
analysis reads the decode off a frozen frame (a runtime that reconstructs the model's decode exactly),
verification proves two-sided packing bounds on the structures (a machine-checked corpus), and learning
holds the decode fixed while \emph{retraining the frame}. We state the calculus, its semiring family,
and the load-bearing theorems---each tagged as machine-proved, empirical, or open---and we cast the
learning problem as frame optimization under explicit soundness invariants.
\end{abstract}

\section{Introduction}

A transformer computes a next-token distribution by projecting a residual vector onto an unembedding
matrix and taking a softmax. Read literally this is a single linear map followed by a normalization; but
the residual vector is itself an \emph{additive} object---a sum of per-layer attention and MLP writes
plus the embedding---so the decode logit of token $v$ decomposes as a \emph{sum of inner products}, one
per contributing block. We take this decomposition as primitive and ask what algebra it lives in. The
answer is a semiring-weighted incidence calculus: tokens are propositions carrying geometric directions,
blocks are sources casting weighted votes, and the decision is the semiring aggregate of those votes.

The payoff of naming the algebra is that three otherwise-separate activities become facets of one
object. \emph{Analysis} of a frozen model is evaluation of the calculus in the log-semiring; the
decomposition is exact, so a runtime can reconstruct the model's own decode argmax with no approximation.
\emph{Verification} is proving structural bounds---how many tokens a frame can decode, how many rules can
write the logits, what interference overpacking forces---and these are theorems about the calculus's
structures, machine-checked in a kernel. \emph{Learning} is the recognition that the frame is a free
parameter: one can hold the decode semantics fixed and deform the geometry so that more structure becomes
retrievable. PIC is the shared vocabulary that lets a definition in one activity be a hypothesis in
another.

This paper defines PIC (\S\ref{sec:core}--\S\ref{sec:calculus}), separates frame from decode
(\S\ref{sec:sep}), collects the proved theorems as a two-sided packing picture (\S\ref{sec:thm}), and
casts frame learning as constrained optimization (\S\ref{sec:learn}). Throughout we tag each claim
\proved{} (machine-checked), \empirical{} (measured), or \open{}.

\section{Core structures}\label{sec:core}

\begin{definition}[Frame and Gram kernel]
Let $V$ be a finite set of \emph{propositions} (tokens), $|V|=n_p$, and let $\HH=\RR^d$ be the residual
space. A \emph{PIC frame} is a map $\Uframe:V\to\HH$, $v\mapsto U_v$. It induces the \emph{Gram kernel}
$\Gram(v,w)=\inner{U_v}{U_w}$ and, after row normalization $\bar U_v=U_v/\lVert U_v\rVert$, the
\emph{cosine Gram} $\rho(v,w)=\inner{\bar U_v}{\bar U_w}$.
\end{definition}

The Gram kernel supplies the continuous, non-truth-functional overlap between propositions: tokens are
not independent atoms but directions with measurable coherence.

\begin{definition}[Sources, incidence, monomial, polynomial]\label{def:agg}
A \emph{source set} $S$ ($|S|=J$) assigns to each source $j$ a contribution $c_j:V\to R$ valued in a
semiring carrier $R$. The \emph{residual reading} fixes a vector $d_j\in\HH$ per source and sets the
\emph{incidence} $c_j(v)=\inner{d_j}{U_v}$. The two semiring operations play different roles:
\emph{$\otimes$ combines sources within a token}, giving a \emph{monomial}
\[
  L(v)=\Bigl(\bigotimes\nolimits_{j\in S} c_j(v)\Bigr)\otimes b_v=\sum_{j\in S}\inner{d_j}{U_v}+b_v
  \qquad(\text{$T$-independent});
\]
while \emph{$\oplus_T$ combines tokens}, giving the decode \emph{polynomial}
\[
  Z_T=\bigoplus\nolimits_{T,\,v\in V} L(v)
      =\bigoplus\nolimits_{T,\,v\in V}\ \bigotimes\nolimits_{j\in S} c_j(v)\otimes b_v .
\]
\end{definition}

Thus each token is a monomial and the decode is a $(\oplus_T,\otimes)$ polynomial whose monomials are the
tokens and whose variables are the sources---the tropical-polynomial picture of a ReLU network. The
decision reads off $Z_T$: at $T\to0$, $Z_0=\max_v L(v)$ and the decode is $\arg\max_v L(v)$; at $T=1$,
$Z_1=\operatorname{logsumexp}_v L(v)$ and $p(v)=\exp(L(v)-Z_1)$. The monomials are shared across $T$; only
$\oplus_T$ changes (Lemma~\ref{lem:tinv}).

In the transformer instance, the sources are the \emph{DLA blocks}: an $L$-layer model has $n_b=2L+1$
blocks (one attention and one MLP/expert write per layer, plus the embedding), and $d_j$ is block $j$'s
additive contribution to the residual at the decode position. The incidence $c_j(v)=\inner{d_j}{U_v}$ is
precisely the per-block contribution a decomposing runtime emits, and its source-sum reconstructs the
model's decode logit.

\paragraph{Native notation.} PIC borrows its carriers (semiring, Gram kernel, Welch bound, power
diagrams) so as to inherit their theorems, but three things are native and get their own notation. The
\emph{incidence arrow} $j\triangleright v:=\inner{d_j}{U_v}$ names the one new primitive---provenance as
an inner product in a learnable frame. The \emph{coalition bracket}
$\llbracket P\rrbracket(v):=\bigotimes_{j\in P}(j\triangleright v)$ is the partial monomial of a coalition
$P\subseteq S$ (no $T$: sources add). The \emph{margin turnstile}
\[
  P\vdash_\gamma v\quad:\Longleftrightarrow\quad
  \llbracket P\rrbracket(v)-\max_{w\neq v}\llbracket P\rrbracket(w)\ \ge\ \gamma
\]
(``coalition $P$ decides $v$ with margin $\ge\gamma$''; write $P\vdash v$ for $\gamma=0$) collapses the
decision regimes to one line each: \emph{retrieved} $\exists j.\ \{j\}\vdash_\gamma v$;
\emph{composed} $S\vdash v$ but $\forall j.\ \neg(\{j\}\vdash v)$; \emph{irreducible} $S\vdash v$ and
$\forall P\subsetneq S,P\neq\emptyset.\ \neg(P\vdash v)$.

\section{The semiring family}\label{sec:semiring}

PIC is parameterized by a temperature $T\ge 0$ selecting a commutative semiring
$(R_T,\oplus_T,\otimes,0,1)$ with a fixed product $\otimes=+$ (incidences along a coalition add) and a
$T$-dependent sum:
\[
  a\oplus_T b \;=\; T\,\log\!\bigl(e^{a/T}+e^{b/T}\bigr),
  \qquad
  a\oplus_T b \xrightarrow[\;T\to 0^+\;]{} \max(a,b).
\]

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
$T$ & semiring & $a\oplus_T b$ & reading \\
\midrule
$1$ & log-semiring & $\log(e^a+e^b)$ & probabilistic incidence $\to$ softmax \\
$\to 0^+$ & tropical (max-plus) & $\max(a,b)$ & argmax, margins, power diagrams \\
general & deformed & $T\log\sum e^{\cdot/T}$ & temperature-controlled decoding \\
\bottomrule
\end{tabular}
\end{center}

The two ends are joined by \emph{Maslov dequantization}: the probabilistic calculus (the model in
operation) and the geometric calculus (the proofs) are limits of one deformation, not different objects.

\begin{definition}[The complete semiring]\label{def:semiring}
For each $T\ge0$, $R_T=(R,\oplus_T,\otimes,\mathbf{0},\mathbf{1})$ has carrier $R=\RR\cup\{-\infty\}$,
sum $a\oplus_T b=T\log(e^{a/T}+e^{b/T})$ with $\oplus_0:=\max$, product $a\otimes b=a+b$, sum identity
$\mathbf{0}=-\infty$, and product identity $\mathbf{1}=0$. It satisfies the semiring axioms: $(R,\oplus_T)$
is a commutative monoid with identity $-\infty$; $(R,\otimes)$ is a commutative monoid with identity $0$;
$\otimes$ distributes over $\oplus_T$; and $\mathbf{0}$ annihilates ($-\infty\otimes a=-\infty$). It is
commutative, and idempotent exactly at $T=0$ ($a\oplus_0 a=a$ but $a\oplus_1 a=a+\log 2$)---the algebraic
difference between geometry and probability.
\end{definition}

\noindent The distributive law $a\otimes(b\oplus c)=(a\otimes b)\oplus(a\otimes c)$ is the load-bearing
one: it rewrites a sum-of-maxes (a ReLU stack) as a max-of-sums (a tropical polynomial). It and the two
monoid laws are machine-checked at $T=0$ \proved{} (\texttt{TropicalSemiring.thy}); for general $T$ the
distributive law is the log-domain identity
$a+T\log(e^{b/T}+e^{c/T})=T\log(e^{(a+b)/T}+e^{(a+c)/T})$.

\section{PIC as a calculus}\label{sec:calculus}

\paragraph{Syntax.} Sources are weighted clauses; a \emph{derivation} of $v$ is a finite
\emph{coalition} $P\subseteq S$ whose incidences combine, under $\otimes$ within the coalition and
$\oplus_T$ across alternatives, to support $v$.

\paragraph{Semantics.} Semiring \emph{provenance} in the style of Green--Karvounarakis--Tannen: the
value of $v$ is an element of $R_T$ accumulated over derivations, with the Gram kernel supplying the
continuous overlap a Boolean provenance lacks. The frozen-model reading is the log-semiring provenance;
the geometric reading is the tropical provenance.

\paragraph{Decision rules} (in the native turnstile of Definition~\ref{def:agg}).
\begin{itemize}[nosep]
  \item \emph{Aggregate:} $L(v)=\llbracket S\rrbracket(v)\otimes b_v$; decode $\arg\max_v L(v)$.
  \item \emph{Retrieved:} $\exists j\in S.\ \{j\}\vdash_\gamma v$ (a singleton decides; high $\mu_t$).
  \item \emph{Composed:} $S\vdash v$ but $\forall j\in S.\ \neg(\{j\}\vdash v)$ (no singleton; $\mu_t=0$).
  \item \emph{Irreducible:} $S\vdash v$ and $\forall P\subsetneq S,P\neq\emptyset.\ \neg(P\vdash v)$.
\end{itemize}

\paragraph{Soundness anchors.}
\begin{itemize}[nosep]
  \item \emph{Log-semiring $\Rightarrow$ transformer} \empirical{}. With inner-product incidences in the
        log-semiring, the calculus is next-token behaviour: a decomposing runtime reconstructs the
        model's decode argmax with measured reconstruction $1.00$ across architectures (rotary, NeoX,
        mixture-of-experts, latent attention). PIC is the forward pass rebracketed, not an approximation.
  \item \emph{Tropical $\Rightarrow$ power diagram.} As $T\to0$, $\arg\max_v \inner{r}{U_v}+b_v$ is
        Laguerre (power-diagram) cell membership over sites $\{U_v\}$ with weights $\{b_v\}$; the
        capacity theorems of \S\ref{sec:thm} are the geometry of that diagram.
\end{itemize}

\noindent The two anchors are consistent only because of a fact native to the $T$-family:

\begin{proposition}[Decode temperature-invariance]\label{lem:tinv}
For every $T>0$, $\arg\max_v L(v)$ is the same token, equal to the tropical ($T\to0$) decoder; only the
normalized confidence $p_T(v)=\exp\bigl((L(v)-Z_T)/T\bigr)$ and the partition $Z_T=\bigoplus_{T,v}L(v)$
depend on $T$.
\end{proposition}
\begin{proof}
The monomials $L(v)=\llbracket S\rrbracket(v)\otimes b_v$ carry no $T$ (sources combine by $\otimes=+$).
$T$ enters only the cross-token aggregator $\oplus_T$, which is strictly order-preserving in each
argument with $\arg\max=(\oplus_0)$-selector; hence the $\arg\max$ of the fixed vector $(L(v))_v$ is
$T$-independent. $\oplus_T$ reweights how much the winner beats the field, never which token wins.
\end{proof}

This is the rigorous bridge between the anchors: the \emph{proofs} run at $T\to0$ (tropical margins,
power-diagram capacity, head/tail) and the \emph{model} runs at $T=1$ (softmax), yet they decode the same
token because they share the monomials. Every margin/decode theorem of \S\ref{sec:thm} proved tropically
therefore transfers verbatim to the operating model's argmax; the $T$-dependent part is exactly the
confidence (entropy, margins), the locus of the forge tax.

\section{Encode, frame, decode}\label{sec:sep}

PIC is a sandwich of three roles: the \emph{encoder} writes the residual, the \emph{frame} is the shared
dictionary, and the \emph{decoder} reads it out. The encode side has an explicit model
(machine-checked, \texttt{PIC\_Core.thy}).

\begin{definition}[Encoder]\label{def:enc}
A PIC encoder fixes a finite rule set $K$, fixed write directions $a:K\to\HH$, and input-dependent gates
$g:K\to(X\to\RR)$. On input $x$ the residual is the gated sum $\operatorname{enc}(x)=\sum_{k\in K}g_k(x)\,a_k$,
and the logit is $L(v;x)=\inner{\operatorname{enc}(x)}{U_v}+b_v=\sum_{k\in K}g_k(x)\inner{a_k}{U_v}+b_v$.
\end{definition}

This expresses any transformer block: a block's residual write is $W_{\mathrm{out}}\cdot(\text{input
vector})$---fixed output columns $a_k$ gated by an input coefficient $g_k(x)$---for MLP and attention
alike. The gate $g_k(x)$ is left \emph{uninterpreted}: PIC formalizes the linear write/read algebra
($\inner{a_k}{U_v}$ is rule $k$'s incidence) and leaves the nonlinear gate as the un-modelled forge tax,
so the encoder sharpens rather than enlarges the calculus. Three facts are proved \proved{}: the forward
pass is one PIC term ($L(v;x)=\sum_k g_k(x)\inner{a_k}{U_v}+b_v$); every encoded residual lies in
$\operatorname{span}\{a_k\}$ of dimension $\le\min(|K|,d)$ (routing rank, now a property of the encoder---
\S\ref{sec:thm}); and every fixed-input slice \emph{is} a source model (Def.~\ref{def:agg}) with
$d_k=g_k(x)a_k$. The remaining two roles split every other quantity:

Every other PIC quantity is either \emph{frame-side} (a function of $\{U_v\}$ alone---intrinsic,
label-free) or \emph{decode-side} (a function of the aggregate $L$ and a target, evaluated in the chosen
semiring).

\paragraph{Decode-side.}
margin $m(t)=L(t)-\max_{v\neq t}L(v)$;\quad
participation ratio $\mathrm{PR}=\bigl(\sum_j|w_j|\bigr)^2/\sum_j w_j^2$ over incidence magnitudes
$w_j=|c_j(t)|$ (the effective number of contributing sources);\quad
multiplicity $\mu_t=\#\{j:\arg\max_v c_j(v)=t\}$ (unanimity; $\mu_t=0$ is composed);\quad
ambiguity $\eta^2$ (between/total variance of candidate firing on the lowest-margin slice).

\begin{remark}[Architecture-fair PR]
When sources carry a common-mode component identical across candidates (e.g.\ a LayerNorm/embedding
offset), it cancels in the argmax but inflates raw $\mathrm{PR}$. The discriminative count centres each
source across candidates first, $\mathrm{PR}\bigl((c_j(t)-\overline{c_j})_j\bigr)$; raw and centred agree
for RMSNorm frames and diverge for LayerNorm. Use the centred form for cross-architecture comparison.
\end{remark}

\paragraph{Frame-side.}
Gram coherence $\max_{v\neq w}|\rho(v,w)|$;\quad
frame potential $\mathrm{FP}=\operatorname{mean}_{v\neq w}\rho(v,w)^2$;\quad
Welch floor $W=(n_p-d)/\bigl(d(n_p-1)\bigr)$ for $n_p>d$, else $0$. Always $\mathrm{FP}\ge W$; the ratio
$\mathrm{FP}/W$ reports frame quality, equal to $1$ at an equiangular tight frame.

\section{The two-sided packing theorems}\label{sec:thm}

The proved results split into a \emph{frame side} (how many tokens a geometry can decode) and a
\emph{generator side} (how many rules can write the logits, and what interference overpacking forces).
This split is not a coincidence: it is the \emph{two arities} of the decode polynomial $Z_T$ of
Definition~\ref{def:agg}. A $(\oplus,\otimes)$ polynomial has a monomial-count and a variable-rank, and
PIC bounds each---the frame side bounds the \textbf{monomials} (tokens that can be cleanly decoded, the
$\oplus$-arity) at $(1+2\rho/\gamma)^d$, the generator side bounds the \textbf{variables} (independent
sources, the $\otimes$-arity) at $\min(M,d)$, and Welch is the obstruction that appears exactly when the
variable side is overpacked. ``Two-sided packing'' is thus a simultaneous bound on a polynomial's
monomial-count and its variable-rank. The frame side is astronomically slack for realistic $d$; the
binding side is the generator---superposition is forced on the writers, not on the readouts.

\subsection{Frame side: decode capacity (monomial count)}

Call $v$ \emph{$\gamma$-decodable} if some unit-ball residual $r$ ($\lVert r\rVert\le1$) makes it win by
margin $\gamma$: $\forall w\neq v,\ \inner{r}{U_w}+b_w+\gamma\le\inner{r}{U_v}+b_v$.

\begin{theorem}[Margin separation \proved{}]\label{thm:sep}
If $v$ and $w$ are both $\gamma$-decodable then $\gamma\le\lVert U_v-U_w\rVert$.
\end{theorem}
\begin{proof}[Proof sketch]
Add the two witness inequalities; the biases cancel, and Cauchy--Schwarz with $\lVert r_v-r_w\rVert\le2$
gives the bound. (Machine-checked in \texttt{DecodeCapacity.thy}.)
\end{proof}

\begin{corollary}[Head capacity \proved{}]\label{cor:cap}
Any set $S$ of $\gamma$-decodable tokens is a $\gamma$-code: $\lVert U_v-U_w\rVert\ge\gamma$ for distinct
$v,w\in S$. Hence $|S|\le(1+2\rho/\gamma)^d$ with $\rho=\max_v\lVert U_v\rVert$.
\end{corollary}

This is the cell-capacity half; the bound is astronomically slack for realistic $d$, so this side rarely
binds.

\subsection{Generator side: routing rank and Welch interference (variable rank)}

$M$ rules write logit adjustments along fixed readout vectors $\{a_1,\dots,a_M\}$ through
input-dependent gates $h_k(z)$.

\begin{theorem}[Routing superposition \proved{}]\label{thm:rank}
$\sum_{i\in I}h_i\,a_i\in\operatorname{span}(a_I)$ and $\dim\operatorname{span}(a_I)\le|I|$. Every rule
adjustment lives in a subspace of dimension $\le\min(M,d)$; superposition is forced when the number of
routing decisions exceeds $M$.
\end{theorem}

\begin{theorem}[Routing Welch \proved{}]\label{thm:welch}
For $n$ unit-norm routing features in $M$ coordinates,
$\sum_{i\neq j}\inner{f_i}{f_j}^2\ge n(n-M)/M$, strictly positive exactly when $n>M$. Cross-talk-free
routing requires $M\ge n$.
\end{theorem}

\begin{remark}[The missing link \open{}]
The step ``routing interference $\Rightarrow$ margin degradation'' is \emph{not} a kernel theorem.
Trained models pack features at $\approx$ the Welch floor with healthy margins; the coherence-to-margin
implication is measured and mild, not proved.
\end{remark}

\subsection{The decode certificate and its robustness}

With $\operatorname{decode}(L,S)=\max_{v\in S}L(v)$:

\begin{theorem}[Head/tail certificate \proved{}]\label{thm:headtail}
$\operatorname{decode}(L,H\cup T)=\max(\operatorname{decode}(L,H),\operatorname{decode}(L,T))$, and if
$\operatorname{decode}(L,T)\le\operatorname{decode}(L,H)$ then the full decode equals the head decode and
its argmax lies in $H$. If instead the head does not dominate, the decision lies in the tail---the
explicit, uncertified residue.
\end{theorem}

\begin{theorem}[Margin certificate, threshold tight \proved{}]\label{thm:margin}
If $|L'(v)-L(v)|\le\delta$ for all $v$ and $m(t)>2\delta$, then $t$ remains the strict argmax under $L'$.
The constant $2\delta$ is tight (a $m(t)=2\delta$ token can be tied by a $\delta$-perturbation), and the
certificate is silent for $m(t)\le2\delta$---exactly the small-margin residue.
\end{theorem}

Theorem~\ref{thm:headtail} is the algebra behind the empirical observation that a per-position sparse
head (a handful of late-layer blocks) reproduces the decode \emph{when} it dominates the tail; harder
tokens push mass into the residue.

\subsection{Irreducibility}

For a composition matrix $c:\text{sources}\times\text{outcomes}\to\RR$, $S$ \emph{decides} $t$ if
$\sum_{j\in S}c(j,v)<\sum_{j\in S}c(j,t)$ for all $v\neq t$; $\mu_0$ holds if no singleton decides $t$;
$S$ is \emph{irreducible} for $t$ if it decides $t$ and no proper nonempty subset does.

\begin{theorem}[$\mu_0\nRightarrow$ irreducible \proved{}]\label{thm:irr}
There is an explicit instance with $\mu_0$ (no single source decides $t$) in which a proper pair already
decides $t$; and a separate explicit fully-irreducible triple. Hence ``no singleton decides'' does not
imply ``all sources needed''.
\end{theorem}

This is the precise statement of $\mu_t=0\neq\text{irreducible}$: low multiplicity is necessary but not
sufficient for genuine irreducibility. Whether a composed token is irreducible under \emph{every}
admissible frame---the real necessity question---is \open{}.

\section{Frame learning as constrained optimization}\label{sec:learn}

Fix the decode semantics (log-semiring $L$, softmax) and optimize the frame $U$---equivalently, deform
the Gram kernel---so that more propositions become retrievable:
\[
  \mathcal{L}(U)=\underbrace{\mathrm{NLL}(L;t)}_{\text{decode, data}}
  +\;\lambda_m\underbrace{\operatorname{ReLU}\!\bigl(\gamma^\star-m(t)\bigr)}_{\text{decode, margin}}
  +\;\lambda_f\underbrace{\mathrm{FP}(U)}_{\text{frame, structure}} .
\]
The decode-side terms anchor the aggregate to the targets and push the worst-competitor margin past
$\gamma^\star$; the frame-side term decorrelates propositions toward the Welch floor. A
Generate--Gate--Refine loop wraps the gradient step: \emph{generate} candidate sources (random or
frame-aligned, or real DLA blocks loaded through the analysis seam), \emph{gate} them by the PIC
diagnostics ($\eta^2$, activation variance, firing decorrelation), \emph{refine} by a step on
$\mathcal{L}$.

\paragraph{Soundness invariants.} The loop should preserve: (S1) \emph{decode soundness}---the refined
frame still realizes a valid log-semiring decoder on held data; (S2) \emph{margin monotonicity}---the
certified margin mass $\#\{t:m(t)>2\delta\}$ of Theorem~\ref{thm:margin} does not decrease; (S3)
\emph{frame admissibility}---$\mathrm{FP}(U)\ge W$ is structural, so a report must target $\mathrm{FP}/W\to1$
and never claim to beat the floor. These are natural targets for learning-side soundness theorems.

\paragraph{What is established \empirical{}.} On synthetic benchmarks no theory-guided knob
(frame-regularization, allocation/targeting, propose-score-select, coherence-regularization) beats plain
SGD on $\mathcal{L}$; PIL training \emph{is} gradient descent on the frame, cheap only as a shallow head
on frozen sources. The win that exists is \emph{retraining the frame/subspace}---the movable side of the
boundary---rather than fitting an encoder into a frozen dictionary, consistent with the verification-side
result that frozen compression hits a $\Theta(d)$ floor a rank-bottlenecked retrained update clears.

\paragraph{The logic-programming reading.} Structurally PIC is a semiring-weighted logic program: the
turnstile $P\vdash_\gamma v$ is a weighted rule (body coalition derives head proposition with margin
$\gamma$), the answer is the least fixpoint of the immediate-consequence operator
$T_P(I)=\{v:\exists P\subseteq I,\ P\vdash v\}$ evaluated in $R_T$, and the query is the decode. The
semiring chooses the logic---$T=0$ is shortest-path/Viterbi (min-cost) logic, $T=1$ is probabilistic
(ProbLog-style) logic, Boolean is classical Datalog---with PIC's one native move that the fact weights are
\emph{derived from geometry}, $j\triangleright v=\inner{d_j}{U_v}$, rather than given. The demand-closure
theorem (\S\ref{sec:thm}, $\operatorname{lfp}(\operatorname{restrict}T\,D)=\operatorname{lfp}T\cap D$
\proved{}) is exactly the soundness of the magic-sets/demand optimization for such a language, and the
margin certificate bounds how far a clause weight may drift before an answer flips. The frame/decode
separation is its type discipline: frame-side clauses (what overlaps what) plus decode-side queries
(margins, multiplicity).

\section{An empirical parameter the proofs do not pin down}\label{sec:tau}

The packing exponent that binds in practice is an effective rank $\tau^\star$, not the ambient $d$. A
natural conjecture, $\tau^\star\approx\min(e^{H},d)$ (effective output support sets the rank), is
\emph{refuted in the cross-model direction}. Measured on a seven-model sweep (Pythia 70m--1.4b, Qwen
0.5B/1.5B) via the analysis seam, the architecture-fair effective decode-block count $r_{\mathrm{eff}}$
\emph{anti-correlates} with $e^{H}$ (Pearson $-0.71$) and \emph{tracks the block count} $n_b=2L+1$ (Pearson
$+0.75$), depth, and width: larger, more confident models use \emph{more} effective blocks, not fewer. The
clean dissociation is that Pythia-1b (16 layers, \emph{more} parameters) uses fewer effective blocks than
Pythia-410m (24 layers, \emph{fewer} parameters), so it is depth/$n_b$---not parameter count and not
entropy---that sets $r_{\mathrm{eff}}\approx(0.4\text{--}0.5)\,n_b$ within an architecture. A genuine but
modest \emph{within-model} effect survives (higher-entropy tokens use somewhat more blocks; Spearman
$\approx0.3$--$0.4$ in capable models). So $\tau^\star$ is real but capacity/depth-bound rather than
entropy-bound; its functional form is \open{}. We record this as the cleanest current example of the
discipline the tags enforce: a plausible law, measured, and reported by what the data support rather than
by what would have been tidy.

\section*{Provenance of claims}
\proved{} statements are machine-checked in the i-orca Isabelle corpus
(\texttt{tropical/\{TropicalSemiring,HeadTail,DecodeCapacity,RoutingRank\}.thy},
\texttt{superposition/\{Welch,RoutingWelch,Superposition\}.thy},
\texttt{provable\_opt/ProvableOpt\_Common.thy}, \texttt{fieldrun/separation/Separation.thy}).
\empirical{} statements are measured by the fieldrun runtime and the pil experiments.
\open{} statements are flagged hypotheses. See \texttt{spec/PIC\_SPEC.md} for the operational
definitions and file-level pointers.

\end{document}
